```{r, echo = FALSE, include = FALSE}
source("./scripts/packages.R") # load necessary packages
df <- read.csv("./data/data_week.csv", header = TRUE, sep = ",", dec = ".", check.names = FALSE)
source("./scripts/boxplot_legend.R")
```


# Building related data

Data are usually described by a set of features also called variables, attributes or dimensions.

- Categorical: Features whose values are taken from a defined set of values.
- Numerical: Features whose values are continuous or integer-valued. They are represented by numbers and possess most of the properties of numbers.

- continuous
- discrete

categorical variables numerical variables

- whole building
- meter level
- system level

- dynamic
- static

## influencing variables
## Time series

## Visualizations

### Box plots {#sec:boxplots}

```{r, hide = TRUE}
p1<- df %>%
  ggplot() +
  geom_histogram(aes(x = Total_Power, y = ..density..),
                 colour = "black", fill = "white", 
                 binwidth = 40, na.rm = T) +
  geom_density(aes(x = Total_Power, y = ..density..),
               alpha = .2, fill = "#FF6666", na.rm = T) + 
  stat_boxplot(aes(x = Total_Power, y = -0.0025), 
               geom ='errorbar', width = 0.0025) +
  geom_boxplot(aes(x = Total_Power, y = -0.0025), 
               fill = "lightgrey", width = 0.0025, na.rm = T) +
  geom_point(aes(x = mean(Total_Power, na.rm = T), y = -0.0025), 
             shape = 3, size = 1.5) +
  labs( x = "Electrical load [kW]" , y = "Density [-]") + 
  theme_classic() + coord_flip()

p2 <- ggplot_box_legend()

ggarrange(p1, p2, nrow = 1)

```

### scatter plots
### carpet plots
### calendar plots

